---
title: "How is the UK healthcare system performing"
author: "Group 13: Amelia Przybyl, Athos Gyalui, Drishti Hoskote, Mingyu Dai, San Kashyap"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: sandstone
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
---



```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, library, include = FALSE}
library(scales)
library(tidyverse) # the usual stuff: dplyr, readr, and other goodies
library(lubridate)
library(ggmap)
library(ggrepel)
library(gridExtra)
library(pander)
library(here)
library(janitor)
library(skimr)
library(leaflet)
library(leaflet.extras)
library(tmap)
library(tmaptools)
library(hrbrthemes)
library(mapview)
library(viridis)
library(plotly)
library(ggtext)
library(GGally) # for correlation-scatter plot matrix
library(treemapify)
library(RColorBrewer)
library(extrafont)
library(ggThemeAssist)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(vroom)
library(patchwork)
library(ggalt)
library(ggplot2)
library(sf)
library(stringr)
library(showtext)
```


```{r, them_function_and_palette, include = FALSE}
theme_nhs <- function() {
  font = "Avenir Next"
  theme_minimal()
  theme(
    panel.grid = element_blank(), 
    panel.background = element_blank(), 
    axis.line = element_line(size = 0.5, colour = "grey80"), 
    # Remove legend
    legend.position = "none", 
    legend.key = element_rect(fill = "transparent"), 
    axis.text.x = element_text(color = "grey20", size = 10, family = font),
    axis.text.y = element_text(color = "grey20", size = 10, family = font),
    axis.title.x = element_text(color = "grey20", size = 12, angle = 0, 
                                hjust = .5, vjust = 0, family = font),
    axis.title.y = element_text(color = "grey20", size = 12, 
                                hjust = .5, vjust = .5, family = font), 
    plot.title = element_text(color = "grey20", size = 14, 
                              family = font, face="bold"), 
    legend.text = element_text(color = "grey20", size = 10, family = font), 
    legend.title = element_text(color = "grey20", size = 12, family = font)
  )
}
palette <- c('#235eb8', '#595dac', '#756f9f', '#8a8191', '#9c9481', '#aba771', '#b9bb5e', '#c6cf45', '#d1e41a')
two_color <- c('#235eb8', '#d1e41a')
```

<!-- setup -->

# Overview of Healthcare Spending

## Healthcare spending over the years

```{r hc_spending_data, warning=FALSE, message=FALSE}

#Current health expenditure (% of GDP) (Source: https://data.worldbank.org/indicator/SH.XPD.CHEX.GD.ZS)
wb_data <- read_csv("data/healthcare-spending/world_bank_data.csv") %>%
  clean_names()

skimr::skim(wb_data)

wb_data <- wb_data %>%
  remove_empty(which=c("rows", "cols"), quiet=FALSE)

# get time series data
wb2 <- wb_data %>%
  pivot_longer(cols=c("x2000", "x2001", "x2002", "x2003", "x2004", "x2005", "x2006", "x2007", "x2008", "x2009", "x2010", "x2011", "x2012", "x2013", "x2014", "x2015", "x2016", "x2017", "x2018", "x2019"),
               names_to="Year",
               values_to="Spending") %>% 
  mutate(Year = as.integer(substr(Year, 2, 5)))
head(wb2)

```


```{r visualisations, warning=FALSE, message=FALSE}

eu <- c("AUT", "GBR", "BEL", "BGR", "CYP", "HRV", "CZE", "DNK", "EST", "FIN", "FRA", "DEU", "GRC", "HUN", "IRL", "ITA", "LVA", "LTU", "LUX", "MLT", "NLD", "POL", "PRT", "ROM", "SVK", "SVN", "ESP", "SWE")

wb2 %>%
  filter(country_code %in% eu) %>%
  mutate(is_gbr = ifelse(country_code=="GBR", 1, 0)) %>%
  group_by(is_gbr, Year) %>%
  summarize(spending_avg=mean(Spending, na.rm=TRUE)) %>%
  mutate(label = ifelse(Year != 2019, NA_character_, ifelse(is_gbr==1, "GBR", "EU"))) %>% 
  ggplot(aes(x=Year, y=spending_avg, color=as.factor(is_gbr))) + 
  geom_vline(
    xintercept = seq(1999, 2019, 1), color = "grey80", size = .4) + 
  geom_segment(
    data = tibble(y = seq(0, 11, 1), x1 = 1999, x2 = 2019),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey80",
    size = .4
  ) +
  geom_segment(
    data = tibble(y = 0, x1 = 1999, x2 = 2019),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey80",
    size = .4
  ) +
  geom_point() +
  geom_line() + 
  labs(title = "<b> UK spends relatively more on healthcare than the EU average</b><br>
       <span style = 'font-size:12pt'>Health Expenditure as % of GDP in <span style='color:#235eb8'> EU countries </span> and <span style='color:#b9bb5e'> the UK </span></span>",
       y = NULL, x = NULL, 
       caption = "Source: The World Bank (https://data.worldbank.org/indicator/SH.XPD.CHEX.GD.ZS)") + 
  theme_nhs() + 
  scale_x_continuous(breaks = seq(1999, 2019, 1), limits = c(1999, 2021), expand = c(0, 0)) + 
  scale_y_continuous(breaks = seq(0, 11, 1), labels = function(x) paste0(x, "%"), limits = c(0, 11)) + 
  theme(axis.text.x = element_text(angle=45, hjust = 1, vjust = 1), 
        axis.ticks = element_line(color = "grey80",),
        plot.title.position = "plot",
        plot.title = ggtext::element_textbox_simple(size=16)) + 
  geom_text_repel(
    aes(color = as.factor(is_gbr), label = label),
    xlim = c(2019, NA),
    family = "Helvetica",
    fontface = "bold",
    size = 4, 
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted"
  ) + 
  scale_color_manual(values = c(palette[1], palette[7]))

```

## Breaking down the cost

```{r}
dataf <- data.frame(
  Year=c("2007/8","2008/9","2009/10","2010/11","2011/12", "2012/13", "2013/14", "2014/15", "2015/16", "2016/17", "2017/18", "2018/19", "2019/20", "2020/21", "2021/22") ,  
  Funding=c(111.4, 116.6 ,124.1 , 124.5, 125.7, 126, 128.5, 131.2, 134.9, 135.7, 138.4, 141.4, 148.9, 191.01, 190.3))

p = ggplot(data= dataf, aes(x=Year, y=Funding, group = 1)) +
    geom_line(linetype= "dashed", color="blue", size=0.5)+
    geom_point(color="red", size=2) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
rect <- data.frame(xmin="2019/20", xmax= "2021/22", ymin=148.9, ymax=Inf)

p + geom_rect(data=rect, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax),
              color="grey20",
              alpha=0.1,
              inherit.aes = FALSE) +
  labs(
   x = "Year",
    y = "Funding (£bn)",
    title = "COVID-19 caused sharp increases in NHS spending",
    subtitle = "NHS Funding by year",
    caption = "Source:The King's Fund analysis of HM Treasury data"
  ) +

annotate("text", x = "2020/21", y = 153, label = "italic(COVID-19)",size = 2.2,
  parse = TRUE) + 
  theme_nhs() + 
  theme(
    panel.grid.major = element_line(color = "grey80"),  
    panel.grid.minor = element_line(color = "grey80")
  )
```

```{r}
datag = data.frame(x = c("Social Protection", "Health", "General Public Services", "Education"), y = c(299, 217, 109, 100))


p <- ggplot(datag, aes(x=x, y=y)) +
  geom_segment(
    aes(x=x, xend=x, y=0, yend=y),
    color=ifelse(datag$x %in% c("Health"), "orange", "grey20"),
    size=ifelse(datag$x %in% c("Health"), 1.3, 0.7)
  ) +
  geom_point(
    color=ifelse(datag$x %in% c("Health"), "orange", "grey20"),
    size=ifelse(datag$x %in% c("Health"), 5, 2)) +
  labs(
    x = "Sector",
    y = "Funding (£bn)",
    title = "Health is the second highest expenditure",
    subtitle = "Government Expenditure by Sector 2021/22",
    caption= "Source:Public spending: a brief introduction-Philip Brien(07 October, 2022)"
  ) +
  coord_flip() + 
  theme_nhs() + 
  theme(
    panel.grid.major = element_line(color = "grey80"),  
    panel.grid.minor = element_line(color = "grey80")
  )
p
```

```{r}
#https://digital.nhs.uk/data-and-information/publications/statistical/acute-patient-level-activity-and-costing/2019-20#resources
cost <- read.csv("data/healthcare-spending/Acute Patient Level Activity and Costing 2019-20 CSV data.csv") %>%
  janitor::clean_names()

cost1 <- cost %>%
  filter(summary_type=="Total Cost",
         org_name=="ALL SUBMITTERS",
         breakdown!="Total Cost",
         breakdown_group_2!="Unknown")

kid <- c("0 years", "01-04 years", "05-09 years", "10-14 years")
young_adult <- c("15-19 years", "20-24 years", "25-29 years")
adult <- c("30-34 years", "35-39 years", "40-44 years", "45-49 years")
elderly <- c("50-54 years", "55-59 years", "60-64 years", "65-69 years")
old <- c("70-74 years", "75-79 years", "80-84 years", "85-89 years", "90-94 years", "95 years or older")

cost2 <- cost1 %>%
  filter(breakdown_group_1 %in% c("Female", "Male")) %>%
  summarize(age=ifelse(breakdown_group_2 %in% kid, "0-14", ifelse(breakdown_group_2 %in% young_adult, "15-29", ifelse(breakdown_group_2 %in% adult, "30-49", ifelse(breakdown_group_2 %in% elderly, "50-69", "70+")))),
            breakdown_group_1,
            value) %>%
  group_by(age, breakdown_group_1) %>%
  summarize(total_cost=sum(as.numeric(value)))

```


```{r}

cost_age <- cost2 %>%
  ungroup %>%
  group_by(age)%>%
  summarize(total_age=sum(as.numeric(total_cost)))

cost_gender <- cost2 %>%
  ungroup %>%
  group_by(breakdown_group_1)%>%
  summarize(total_gender=sum(as.numeric(total_cost)))

source <- c(cost_age$age, rep("Cost", times = nrow(cost_gender)))
target <- c(rep("Cost", times = nrow(cost_age)), cost_gender$breakdown_group_1)

df_sanky <- data.frame(
  source = source,
  target = target,
  value = c(cost_age$total_age, cost_gender$total_gender)
)

```

```{r snakey}
library(networkD3)

nodes <- data.frame(
  name=c(as.character(df_sanky$source), 
  as.character(df_sanky$target)) %>% unique()
)

df_sanky$IDsource <- match(df_sanky$source, nodes$name)-1 
df_sanky$IDtarget <- match(df_sanky$target, nodes$name)-1

node_color <- 'd3.scaleOrdinal() .domain(["0-14", "15-29", "30-49", "50-69", "70+", 
"Cost", "Female", "Male"]) .range(["#235eb8", "#235eb8", "#756f9f" , "#8a8191", "#9c9481", "#aba771", "#b9bb5e", "#c6cf45"])'

p1 <- sankeyNetwork(Links = df_sanky, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name", 
              colourScale = node_color,
              sinksRight=FALSE) 


p1


```


```{r 1}

#how do costs compare?
cost3 <- cost %>%
  filter(org_name=="ALL SUBMITTERS",
         breakdown!="Total Cost",
         breakdown=="Diagnosis") %>%
  pivot_wider(names_from=summary_type,
              values_from=value) %>%
  janitor::clean_names() %>%
  mutate(cost_per_visit = as.numeric(total_cost)/as.numeric(activity_count)) %>%
  arrange(#desc(activity_count), 
          desc(cost_per_visit))

chosen_diseases <- c("Pregnancy, childbirth and the puerperium", "Cardiac conditions", "Poisoning (inc overdose)", "Diabetes and other endocrinological conditions", "Diseases of the respiratory system", "Bites/stings", "Infectious disease", "Diseases of the digestive system", "Mental and behavioural disorders")

cost3 %>%
  filter(breakdown_group_1 %in% chosen_diseases) %>%
  ggplot(aes(x=fct_reorder(breakdown_group_1, as.numeric(cost_per_visit)), y=cost_per_visit, fill=breakdown_group_1)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme_nhs() + 
  scale_fill_manual(values=c(palette)) + 
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  labs(title = "Costs per visit vary significantly",
       subtitle="Average cost per visit by category",
       caption="Source: Digital NHS, Acute Patient Level Activity and Costing, 2019-20")
```


# Workforce of Healthcare

## Healthcare workforce in the history


```{r, warning=FALSE, message=FALSE}
nhs_earnings<-read_csv("data/healthcare-spending/NHS Staff Annual Earnings Estimates to June 2022 in NHS Trusts and other core organisations in England, Provisional Statistics CSV text file.csv") %>%
  clean_names()

nhs_earnings<-nhs_earnings%>%
  mutate(year=format(date, format="%Y"))%>%
  mutate(month=format(date, format="%m"))%>%
  filter(!is.na(sample_size))%>%
  mutate(gov_spending=amount*sample_size)%>%
  mutate(year=format(date, format="%Y"))%>%
  mutate(month=format(date, format="%m"))

nhs_earnings_sep<-nhs_earnings%>%
  filter(month=="08")%>%
  filter(payment_type=="PUBGRP_010_BASIC_PAY_PER_FTE")%>%
  filter(staff_group=="All staff")
```

```{r}
countperyear<-nhs_earnings%>%
  filter(payment_type=="PUBGRP_010_BASIC_PAY_PER_FTE")%>%
  filter(staff_group=="All staff")


nhs_earnings_sep<-nhs_earnings%>%
  filter(month=="08")%>%
  filter(payment_type=="PUBGRP_010_BASIC_PAY_PER_FTE")%>%
  filter(staff_group=="All staff")



g<-ggplot(countperyear,aes(x=date, y=sample_size/1000000))+geom_line()+
  labs(x="Date",y="Number of Workers (million)",title="Steady increase in NHS workers since 2013",subtitle = "Number of emplyees over time", caption = "NHS Workforce Statistics - July 2022 (NHS, 2022)")+
  geom_point(data=nhs_earnings_sep, aes(x=date, y=sample_size/1000000), colour="#F8766D") + 
  theme_nhs()
g
```


## Staff Shortages

```{r}
vacancy <- read.csv("data/vacancy/vacancy_region.csv") %>% 
  clean_names() %>% 
  rename(nhscr16nm = region) %>% 
  mutate(vacancy = as.numeric(str_remove(vacancy, "%")), 
         vacancy_number = as.numeric(str_remove_all(vacancy_number, ",")))
vacancy
```


```{r, map of UK}
# make sure you have the same direcory stucture to get London wards shapefile
uk_map <- read_sf("data/vacancy/NHS_shape/NHS_Commissioning_Regions_April_2016_Generalised_Clipped_Boundaries_in_England.shp")

uk_map_wgs84 <-  uk_map %>%
  st_transform(4326)
```

```{r}
uk_map_wgs84_with_value <- uk_map_wgs84 %>% 
  mutate(nhscr16nm = ifelse(nhscr16nm == "North Of England", "North of England", nhscr16nm)) %>%
  left_join(vacancy, by = "nhscr16nm")
```

```{r, message=FALSE, warning=FALSE}
vac_map <- uk_map_wgs84_with_value %>% 
  ggplot(aes()) +
  geom_sf(aes(fill = vacancy, text = paste(nhscr16nm, "\nVacancy rate: ", vacancy, "%", sep = "")), colour = "#ffffff", size = 0.1) + 
  scale_fill_gradientn("%Vacancy", colors = c("#153A70", "#235eb8", "#296ED6", "#3082FC"), 
                       labels = c("10%", "9%", "8%", "7%"), 
                       breaks = c(0.1, 0.09, 0.08, 0.07)) + 
  theme_nhs() + 
  theme(
    legend.position = "right"
  ) + 
  labs(title = "London NHS Vacancy was above 10% \n Source: NHS Digital, 2021/22 Q4 (Mar-22)")
ggplotly(vac_map, tooltip = "text")
```


```{r, read_physicians}
physicians <- read_csv('data/vacancy/physicians_per_1000.csv', show_col_types = FALSE)
```


```{r clean_and_extract, warning=FALSE, message=FALSE}

physicians_clean <- physicians %>%
  select("Country Name", "2010", "2019") %>%
  pivot_longer(cols = 2:3, names_to = "year", values_to = "physicians_per_1000") %>%
  mutate(year = as.Date(paste(year, 1, 1, sep = "-")),
         year = year(ymd(year))) %>%
  clean_names()
  

p1 <- physicians_clean %>%
  filter(year == 2019) %>%
  drop_na() %>%
  arrange(desc(physicians_per_1000)) %>%
  top_n(10) %>%
  mutate(color = as.factor(ifelse(country_name == "United Kingdom", 1, 0))) %>% 
  ggplot(aes(x = physicians_per_1000, y = fct_reorder(country_name, physicians_per_1000, max, desc = TRUE))) +
  geom_col(aes(fill = color)) + 
  scale_fill_manual(values = two_color) + 
  theme_nhs() + 
  labs(y = NULL, x = 2019) + 
  theme(
    panel.grid.major.x = element_line(color = "grey80"),  
    panel.grid.minor.x = element_line(color = "grey80")
  )

p2 <- physicians_clean %>%
  filter(year == 2010) %>%
  drop_na() %>%
  arrange(desc(physicians_per_1000)) %>%
  top_n(10) %>%
  mutate(color = as.factor(ifelse(country_name == "United Kingdom", 1, 0))) %>% 
  ggplot(aes(x = physicians_per_1000, y = fct_reorder(country_name, physicians_per_1000, max, desc = TRUE))) +
  geom_col(aes(fill = color)) + 
  scale_fill_manual(values = two_color) + 
  theme_nhs() + 
  labs(y = NULL, x = 2010) + 
  theme(
    panel.grid.major.x = element_line(color = "grey80"),  
    panel.grid.minor.x = element_line(color = "grey80")
  )

(p2 + p1) + 
  plot_annotation(
    title = "Physicians per 1k population increased from 2010 to 2019", 
    subtitle = "Year 2010 v.s. Year 2019", 
    caption = "NHS Vacancy Statistics England April 2015 – June 2022 Experimental Statistics"
  ) &
  theme_nhs() 
```

## How has the NHS salary changed given inflation?

```{r, Merge Average weekly earnings, warning=FALSE, message=FALSE}

# NHS Staff Earnings Estimates (Source: https://digital.nhs.uk/data-and-information/publications/statistical/nhs-staff-earnings-estimates)
nhs_salary <- read_csv("data/healthcare-spending/NHS Staff Annual Earnings Estimates to June 2022 in NHS Trusts and other core organisations in England, Provisional Statistics CSV text file.csv") %>%
  clean_names()

#Average weekly earnings of people in the UK Whole Economy Level (£): Seasonally Adjusted Total Pay Excluding Arrears (Source:https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/earningsandworkinghours/timeseries/kab9/emp)

uk_salary <- read_csv("data/healthcare-spending/UK_salary.csv", skip=7)%>%
  clean_names() %>%
  filter(important_notes %in% c("2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021")) %>%
  mutate(yearly_salary_uk=x2*52) %>%
  rename(year=important_notes) %>%
  summarize(year, yearly_salary_uk)

nhs_salary2 <- nhs_salary %>%
  filter(payment_type == "PUBGRP_020_EARNINGS",
         staff_group == "All staff") %>%
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarize(yearly_earnings=sum(amount)) %>%
  filter(year != "2022") %>%
  rename(yearly_earnings_nhs=yearly_earnings)

salary_data <- merge(uk_salary, nhs_salary2, by="year")

salary_data1 <- pivot_longer(salary_data,
             cols=2:3,
             names_to="type",
             values_to="yearly_salary")
```

```{r, nhs_v_avg_plot}
salary_data1 %>%
  ggplot(aes(x=year, y=yearly_salary, color=type)) +
  geom_point() + 
  theme_nhs() + 
  theme(legend.position = "none", 
        panel.grid.major = element_line(color = "grey80"),  
        panel.grid.minor = element_line(color = "grey80"),
        plot.title.position = "plot",
        plot.title = ggtext::element_textbox_simple(size=16)) + 
  scale_color_manual(values = c(palette[1], palette[7])) +
  labs(title = "<b> NHS salaries are bad... but maybe it's just a sign of a wider problem?</b><br>
       <span style = 'font-size:12pt'> Average salary (£) over the years in the <span style='color:#235eb8'>NHS</span> compared to the <span style='color:#b9bb5e'>UK average </span></span>",
       y = NULL, x = NULL, 
       caption = "Source: ONS Whole Economy Level (£): Seasonally Adjusted Total Pay, NHS Staff Earning Estimates")

```




# Patient Experience


## Data cleaning

```{r, load_clean_data_wl, warning=FALSE, message=FALSE}
# load data
wl <- read.csv("data/waiting_list/nhs_waiting_list_time_series.csv")

wl_ts <-wl %>% 
  pivot_longer(-month, names_to = "indicator", values_to = "value") %>% 
  separate(indicator, c("category", "index"), "_") %>% 
  mutate(category = str_replace_all(category, "\\.", "_"), 
         index = str_replace_all(str_replace_all(index, "\\.", "_"), 
                                 "__", "_")) %>% 
  mutate(index = if_else(str_detect(index, "_$"), 
                         substr(index, 1, nchar(index) - 1), index), 
         category = if_else(str_detect(category, "_$"), 
                            substr(category, 1, nchar(category) - 1), 
                            category)) %>% 
  mutate(index = if_else(str_detect(index, "No__"), 
                         str_replace(index, "No__", "Num_more_"), 
                         str_replace(index, "No", "Num"))) %>% 
  mutate(index = if_else(str_detect(index, "^_"), 
                         str_replace(
                           str_replace(index, "_within", "perc_within"), 
                           "__", "perc_"), 
                         index)) %>% 
  mutate(month = as.Date(paste("01-", substr(month, 4, 6), 
                               "-", substr(month, 1, 2), sep=""), format = "%d-%b-%y")) %>% 
  mutate(category = as.factor(category), 
         index = as.factor(index))
# show data
head(wl_ts)
skimr::skim(wl_ts)
```

## The Expending Waiting List {.tabset .tabset-fade}

Since the pandemic, the number of patients waiting for treatment has been increasing. 

### Waiting List Expansion

The waiting list's size has accually doubled compared with that 6 years ago. 

```{r, plot_wl_incomplete_RTT3, message=FALSE, warning=FALSE}
data <- wl_ts %>% 
  filter(category == "Incomplete_RTT_pathways" &
         index %in% c("Num_within_18_weeks", "Num_more_18_weeks") & 
           month >= "2012-09-01") %>% 
  group_by(month) %>% 
  summarize(total_wl = sum(value))

p_wl_incomplete_rtt <- data %>% 
  ggplot(aes(x = month, y = total_wl)) + 
  theme_nhs() + 
  theme(panel.grid.major.y = element_line(color = "grey80"), 
        panel.grid.minor.y = element_line(color = "grey80"), 
        axis.text.x = element_text(angle=45, hjust = 1, vjust = 1), 
        axis.ticks = element_line(color = "grey80")) + 
  geom_bar(stat = "identity", fill = two_color[1]) + 
  scale_y_continuous(expand = c(0, 0), 
                     labels = unit_format(unit = "million", scale = 1e-6), 
                     breaks = seq(0, 7e6, 1e6)) + 
  scale_x_date(limits = as.Date(c("2012-09-01", "2022-09-01")), 
               expand = c(0, 0), date_labels = "%b-%y", 
             breaks = date_breaks(width = "6 month")) + 
  labs(x = NULL, y = NULL, 
       title = "NHS waiting list has been increasing since the pandemic", 
       subtitle = "Sep 2012 to Sep 2022", 
       caption = "Source: Consultant-led Referral to Treatment Waiting Times Data 2022-23"
       ) + 
  geom_vline(xintercept=as.numeric(as.Date("2020-03-01")), linetype=2, col = "black") + 
  annotate("text", x = as.Date("2018-09-01"), y = 6.5*1e6, 
           label = "Start of the pandemic", hjust = 0.5, color = "black")
p_wl_incomplete_rtt
```

### Forever Waiting...

```{r, plot_wl_incomplete_RTT2, message=FALSE, warning=FALSE}
data <- wl_ts %>% 
  filter(category == "Incomplete_RTT_pathways" &
         index %in% c("Num_more_18_weeks", "Num_more_52_weeks") & 
           month >= "2012-09-01") %>% 
  group_by(month, index) %>% 
  summarize(total_wl = sum(value)) %>% 
  mutate(label = ifelse(month != as.Date("2022-09-01"), NA_character_, 
                        ifelse(index == "Num_more_18_weeks", "18+ w", "52+ w")))

p_wl_18_52 <- data %>% 
  ggplot(aes(x = month, y = total_wl, fill = as.factor(index))) + 
  theme_nhs() + 
  theme(axis.line.x = element_blank(), 
        axis.text.x = element_text(angle=45, hjust = 1, vjust = 1), 
        axis.ticks = element_line(color = "grey80")) + 
  geom_segment(
    data = tibble(y = seq(0, 3200000, 400000), 
                  x1 = as.Date("2012-09-01"), 
                  x2 = as.Date("2022-09-01")),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey80",
    size = .4
  ) +
  geom_segment(
    data = tibble(y = 0, x1 = as.Date("2012-09-01"), x2 = as.Date("2022-09-01")),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey40",
    size = .4
  ) + 
  geom_area(stat = "identity", position = "identity", color = two_color[1], alpha = 0.5) + 
  scale_fill_manual(values = c("#7393E6", palette[1])) + 
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 3200000, 400000), 
                     label = comma) + 
  labs(x = NULL, y = NULL, 
       title = "Number of patients waiting over 18 and 52 weeks", 
       subtitle = "Sep 2012 to Sep 2022", 
       caption = "Source: Consultant-led Referral to Treatment Waiting Times Data 2022-23"
       ) + 
  geom_vline(xintercept=as.numeric(as.Date("2020-03-01")), linetype=2, col = "black") + 
  annotate("text", x = as.Date("2018-09-01"), y = 3000000, 
           label = "Start of the pandemic", hjust = 0.5, color = "black") + 
  scale_x_date(limits = c(as.Date("2012-09-01"), as.Date("2024-03-01")), 
               date_labels = "%y-%b", expand = c(0, 0), 
               breaks = c(as.Date("2012-09-01"), as.Date("2013-03-01"), 
                          as.Date("2013-09-01"), as.Date("2014-03-01"), 
                          as.Date("2014-09-01"), as.Date("2015-03-01"), 
                          as.Date("2015-09-01"), as.Date("2016-03-01"), 
                          as.Date("2016-09-01"), as.Date("2017-03-01"), 
                          as.Date("2017-09-01"), as.Date("2018-03-01"), 
                          as.Date("2018-09-01"), as.Date("2019-03-01"), 
                          as.Date("2019-09-01"), as.Date("2020-03-01"), 
                          as.Date("2020-09-01"), as.Date("2021-03-01"), 
                          as.Date("2021-09-01"), as.Date("2022-03-01"), as.Date("2022-09-01")
                          ))+
  geom_text_repel(
    aes(color = index, label = label),
    xlim = c(as.Date("2022-09-01"), NA),
    family = "Helvetica",
    fontface = "bold",
    size = 4, 
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted"
  ) + 
  scale_color_manual(values = c("#7393E6", palette[1]))
p_wl_18_52
```

### %orever waiting

```{r, Percentage of forever waiting, message=FALSE, warning=FALSE}
data <- wl_ts %>% 
  filter((category == "Incomplete_RTT_pathways") &
         (index %in% c("perc_within_18_weeks", "perc_52_weeks")) & 
           (month >= "2012-09-01")) %>% 
  group_by(month, index) %>% 
  summarize(total_wl = sum(value)) %>% 
  mutate(total_wl = ifelse(index == "perc_within_18_weeks", 1 - total_wl, total_wl), 
         index = ifelse(index == "perc_within_18_weeks", "perc_above_18_weeks", "perc_52_weeks")) %>% 
  mutate(label = ifelse(month != as.Date("2022-09-01"), NA_character_, 
                        ifelse(index == "perc_above_18_weeks", "%18+ w", "%52+ w")))

p_wl_18_52_perc <- data %>% 
  ggplot(aes(x = month, y = total_wl, fill = as.factor(index))) + 
  theme_nhs() + 
  theme(axis.line.x = element_blank(), 
        axis.text.x = element_text(angle=45, hjust = 1, vjust = 1), 
        axis.ticks = element_line(color = "grey80")) + 
  geom_segment(
    data = tibble(y = seq(0, 0.6, 0.05), 
                  x1 = as.Date("2012-09-01"), 
                  x2 = as.Date("2022-09-01")),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey80",
    size = .4
  ) +
  geom_segment(
    data = tibble(y = 0, x1 = as.Date("2012-09-01"), x2 = as.Date("2022-09-01")),
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE,
    color = "grey40",
    size = .4
  ) + 
  geom_area(stat = "identity", position = "identity", color = two_color[1], alpha = 0.5) + 
  scale_fill_manual(values = c(palette[1], "#7393E6")) + 
  scale_y_continuous(labels = scales::percent, 
                     expand = c(0, 0), breaks = seq(0, 0.6, 0.05)) + 
  labs(x = NULL, y = NULL, 
       title = "Percentage of patients waiting over 18 and 52 weeks", 
       subtitle = "Sep 2012 to Sep 2022", 
       caption = "Source: Consultant-led Referral to Treatment Waiting Times Data 2022-23"
       ) + 
  geom_vline(xintercept=as.numeric(as.Date("2020-03-01")), linetype=2, col = "black") + 
  annotate("text", x = as.Date("2018-09-01"), y = 0.57, 
           label = "Start of the pandemic", hjust = 0.5, color = "black") + 
  scale_x_date(limits = c(as.Date("2012-09-01"), as.Date("2024-03-01")), 
               date_labels = "%y-%b", expand = c(0, 0), 
               breaks = c(as.Date("2012-09-01"), as.Date("2013-03-01"), 
                          as.Date("2013-09-01"), as.Date("2014-03-01"), 
                          as.Date("2014-09-01"), as.Date("2015-03-01"), 
                          as.Date("2015-09-01"), as.Date("2016-03-01"), 
                          as.Date("2016-09-01"), as.Date("2017-03-01"), 
                          as.Date("2017-09-01"), as.Date("2018-03-01"), 
                          as.Date("2018-09-01"), as.Date("2019-03-01"), 
                          as.Date("2019-09-01"), as.Date("2020-03-01"), 
                          as.Date("2020-09-01"), as.Date("2021-03-01"), 
                          as.Date("2021-09-01"), as.Date("2022-03-01"), as.Date("2022-09-01")
                          ))+
  geom_text_repel(
    aes(color = index, label = label),
    xlim = c(as.Date("2022-09-01"), NA),
    family = "Helvetica",
    fontface = "bold",
    size = 4, 
    segment.size = .7,
    segment.alpha = .5,
    segment.linetype = "dotted"
  ) + 
  scale_color_manual(values = c(palette[1], "#7393E6"))
p_wl_18_52_perc
```

### Breakdown

```{r}
bd <- read.csv("data/waiting_list/breakdown.csv")
```

```{r}
names(bd) <- c("function_code", "function_name", "month", "total_wl", "total_within_18w", "perc_within_18w", "total_52plusw")
```


```{r}
bd_clean <- bd %>% 
  mutate(
    month = as.Date(paste("01-", substr(month, 4, 6), 
                               "-", substr(month, 1, 2), sep=""), format = "%d-%b-%y")
  )
```


```{r}
total_22 <- (bd_clean %>% filter(month == as.Date("2022-09-01")) %>% select(function_code, total_wl) %>% rename("Sep-2022" = "total_wl"))
total_19 <- (bd_clean %>% filter(month == as.Date("2019-09-01")) %>% select(function_code, total_wl) %>% rename("Sep-2019" = "total_wl"))
total_2yr <- total_19 %>% 
  left_join(total_22, by = "function_code") %>% 
  left_join(
  bd_clean %>% 
    filter(month == as.Date("2022-09-01")) %>% 
    select(function_code, function_name), by = "function_code"
) %>% 
  select(function_name, `Sep-2019`, `Sep-2022`)
```


```{r}
total_2yr_4plot <- total_2yr %>% 
  rowwise() %>% 
  mutate( mean_tot = mean(c(`Sep-2019`,`Sep-2022`) )) %>% 
  arrange(mean_tot) %>% 
  mutate(function_name=factor(function_name, function_name))
 
# Plot
total_2yr_4plot %>% filter(function_name != "Other") %>% 
ggplot() +
  geom_segment( aes(x=function_name, xend=function_name, 
                    y=`Sep-2019`, yend=`Sep-2022`), color="grey20") +
  geom_point( aes(x=function_name, y=`Sep-2019`), color=two_color[1], size=3 ) +
  geom_point( aes(x=function_name, y=`Sep-2022`), color=two_color[2], size=3 ) +
  coord_flip()+
  theme_nhs() +
  theme(
    legend.position = "none",
    panel.grid.major = element_line(color = "grey80")
  ) +
  scale_y_continuous(label = comma) + 
  labs(
    title = "Waiting List Expansion by service", 
    subtitle = "Sep 2019 v.s. Sep 2022", 
    y = "Total Number of Incomplete Pathways", x = "")
```

# Reference

<ol>

<li>[The NHS has a lot of locums. It should listen to them. ](https://www.economist.com/britain/2022/09/22/the-nhs-has-a-lot-of-locums-it-should-listen-to-them)</li>

<li>
[Health at a Glance 2021: OECD Indicators, OECD Publishing, Paris.](https://www.oecd-ilibrary.org/docserver/ae3016b9-en.pdf?expires=1668858641&id=id&accname=guest&checksum=9EED8139CB133279FCC5822C30D734C3)
</li>
[Papanicolas I, Mossialos E, Gundersen A, Woskie L, Jha A K. Performance of UK National Health Service compared with other high income countries: observational study, BMJ, 2019.](https://www.bmj.com/content/367/bmj.l6326)
</ol>
